An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-07-15 EconGrowth.Rnw
Moved the location of the scatter plots for CES models with energy to be AFTER the partial residual plots.
Edited the text of the CDe section. Edited the to-do list. Changed names of resample plots for CES to be consistent with SF, CD, and LINEX.

\section*{To-do List}
\begin{itemize}
\item{Pruim: Add base fit red dots to all cloud plots.}
      \begin{itemize}
      \item{SF}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{CES}
      \item{CESe}
      \item{LINEX}
      \end{itemize}
\item{Heun: Figure out why CESe for ZA with $x$ fails during resampling.}
\item{Heun: Review all assumptions for CES with energy models. 
There are several assumptions for rho and rho1 for various countries. 
When we are resampling, we probably don't want to specify those values.}
\item{Heun: Investigate possible conflicting use of exergy by Warr and Ayres.}
\item{Heun: Fix the energy sankey diagram from the seminar}
\item{Heun: Include adjusted sankey diagram in the seminar}
\item{Heun: Add CES production function fits}
      \begin{itemize}
      \item{Write up the decisions about parameter estimation for the 
            CES production function fits}
      \item{Revisit my assumptions for the CES production function 
            (with energy) for each country.}
      \item{Fix the problems with $\sigma_1$ and $\sigma$ values in 
            the CES parameter tables in the appendix}
      \item{Add CES rows to the AIC table. Code is in the 
            \texttt{createAICTable} function, but it is not yet working. 
            Need to find a way to calculate AIC for the object that is 
            returned from the \texttt{cesEst} function.}
      \end{itemize}
\item{Heun: Fix warnings of the form \texttt{Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562} in the code that generates the Cobb-Douglas with energy fits.}
\item{de Wit: review, critique, and contextualize.}
\item{\st{Heun: Add option for $m = 1$ in single-factor models.}}
\item{\st{Heun: Re-implement confidence interval calculations to use
resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Include scatter graphs for model parameters 
estimated by resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Decide if the single factor models should include constraints
on $\alpha$, $\beta$, and $\gamma$. At present, the single-factor model allows
values outside the ranges $\alpha \in (0,1)$, $\beta \in (0,1)$, and $\gamma \in (0,1)$.}}
\item{\st{Heun: Make smaller scatter plots for resample coefficients for those 
countries where $u$ is involved.}}
\item{\st{Heun: Add a table that shows the fraction of unconverged resample fits
for each model, energy type, and factor.}}
\item{\st{Heun: Get AIC calculations working again.}}
\end{itemize}

@@

\noindent Equation~\ref{eq:CES} is a CES production function with 
capital stock ($k$) and labor ($l$) factors of production. 
Equation~\ref{eq:CESe} augments Equation~\ref{eq:CES} with energy 
using a ($kl$)($e$) nesting structure, as is typical in the literature. 
Equation~\ref{eq:CES} is a degenerate form of Equation~\ref{eq:CESe} 
where $\delta = 1$. The elasticity of substitution between capital ($k$) 
and labor ($l$) is given by $\sigma_1 = \frac{1}{1+\rho_1}$, 
the elasticity of substitution between ($kl$) and ($e$) is given by 
$\sigma = \frac{1}{1+\rho}$, $\delta \in [0, 1]$ describes the importance 
of ($kl$) relative to ($e$), and $\gamma$ is a fitting parameter, 
expected to be close to unity. As $\sigma \rightarrow \infty$, 
($kl$) and ($e$) are perfect substitutes. As $\sigma \rightarrow 0$, 
($kl$) and ($e$) are perfect complements: no substitution is possible.

@@

\noindent According to the LINEX model, an economy that can increase 
capital stock ($k$) without a commensurate increase in labor ($l$) 
and energy ($e$) will experience an increase in output ($y$) in excess 
of its increase of energy consumption ($e$), because $A$ will be 
greater than 1. Similarly, an economy benefits by increasing labor ($l$) 
without a commensurate increase in energy ($e$). Thus, $\rho_k$ is an 
efficiency of using additional labor and energy to make 
additional capital stock available to the economy, 
and $\rho_l$ is an efficiency of using additional energy to make 
additional labor available to the economy. 
$a_0$ represents the overall importance of these efficiencies, 
and $c_t$ represents the relative importance 
of the labor efficiency term compared to the capital stock efficiency term. 
We treat $a_0$ and $c_t$ as fitting parameters; 
both are assumed constant with respect to time.

@@

\noindent to ensure $\alpha + \beta = 1$ for constant returns to scale. 
By constraining $\alpha \in [0,1]$, we ensure that $\beta \in [0,1]$ also. 
**** Is this truly how we implemented the constraints? 
Finish after we actually implement the constraints. **** 
$\lambda$ and $\alpha$ were estimated by the \citet{R} function \texttt{nls}. 
The value of $\beta$ can be found with $\beta = 1 - \alpha$. 
Figure~\ref{fig:CD_Params_Graph} shows values of and 95\% confidence intervals 
for $\lambda$ and $\alpha$. Appendix~\ref{sec:Model_Parameters} 
contains tables with numerical values of model parameters and confidence intervals.

@@

Converting the resample data into 95\% confidence intervals 
provides data for Figure~\ref{fig:CD_Params_Graph}.

@@

\begin{equation} \label{eq:CDe_reparam}
  y = A k^{\min(a,b)}l^{|b-a|}e^{1 - \max(a,b)}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent to ensure that $\alpha + \beta + \gamma = 1$. 
By constraining $a \in [0,1]$ and $b \in [0,1]$, 
we ensure that $\alpha \in [0,1]$, $\beta \in [0,1]$, and $\gamma \in [0,1]$. 
Given values of $a$ and $b$ that represent the least-squares parameter 
optimization for a given economy, 
$\alpha = \min(a,b)$, 
$\beta = |b-a|$, and 
$\gamma = \max(a,b)$ 
can be calculated. 
The \citet{R} function \texttt{nls} was used to estimate the 
least-squares values of $\lambda$, $a$, and $b$.

The Cobb-Douglas with energy parameter estimation process is challenging. 
To demonstrate the characteristics of the parameter estimation process, 
we performed a grid search over values of $\gamma$, 
wherein we re-estimated values of $\lambda$, $\alpha$, and 
$\beta$ at each value of $\gamma$. 
Figure~\ref{fig:CDeGridLattice} shows the sum of squared errors ($SSE$), 
$\alpha$, and $\beta$ as a function of the fixed value of 
$\gamma$ with thermal energy ($q$). 
For most economies, the minimum $SSE$ occurs at a boundary of $\alpha$, 
$\beta$, or $\gamma$. 
The parameter estimation algorithm has difficulty assessing confidence 
intervals for $\alpha$, $\beta$, and $\gamma$, because the constrained 
model behaves very differently on either side of the constraint.

**** Fix the right y axis label. SSE should be italicized. 
Also, scaling on the axis labels and scales is not being picked up. ****

@@

Figure~\ref{fig:CDq_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ parameter values for resample models for the Cobb-Douglas model (with $q$).

@@

<<CES_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesLambdaGammaResamplePlot(energyType=NA)
@

<<CES_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesSigma_1Delta_1ResamplePlot(energyType=NA)
@

@@

<<CES_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\rho$ and $\\delta$.">>=
cesSigmaDeltaResamplePlot(energyType=NA)
@

<<CESq_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesLambdaGammaResamplePlot(energyType="Q")
@

<<CESq_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesSigma_1Delta_1ResamplePlot(energyType="Q")
@

<<CESq_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesSigmaDeltaResamplePlot(energyType="Q")
@

<<CESx_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesLambdaGammaResamplePlot(energyType="X")
@

<<CESx_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesSigma_1Delta_1ResamplePlot(energyType="X")
@

<<CESx_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesSigmaDeltaResamplePlot(energyType="X")
@

<<CESu_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesLambdaGammaResamplePlot(energyType="U")
@

<<CESu_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesSigma_1Delta_1ResamplePlot(energyType="U")
@

<<CESu_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesSigmaDeltaResamplePlot(energyType="U")
@

@@

>>>>

\section*{To-do List}
\begin{itemize}
\itme{Heun: Look into constraining $\delta \in [0,1]$
and $\delta_1 \in [0,1]$}
\item{Pruim: Respond to Heun's wish list for graphs.}
\item{Heun: Figure out why CESe for ZA with $x$ fails during resampling.}
\item{Heun: Review all assumptions for CES with energy models. 
There are several assumptions for rho and rho1 for various countries. 
When we are resampling, we probably don't want to specify those values.}
\item{Heun: Investigate possible conflicting use of exergy by Warr and Ayres.}
\item{Heun: Fix the energy sankey diagram from the seminar}
\item{Heun: Include adjusted sankey diagram in the seminar}
\item{Heun: Add CES production function fits}
      \begin{itemize}
      \item{Write up the decisions about parameter estimation for the 
            CES production function fits}
      \item{Revisit my assumptions for the CES production function 
            (with energy) for each country.}
      \item{Fix the problems with $\sigma_1$ and $\sigma$ values in 
            the CES parameter tables in the appendix}
      \item{Add CES rows to the AIC table. Code is in the 
            \texttt{createAICTable} function, but it is not yet working. 
            Need to find a way to calculate AIC for the object that is 
            returned from the \texttt{cesEst} function.}
      \end{itemize}
\item{Heun: Fix warnings of the form \texttt{Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562} in the code that generates the Cobb-Douglas with energy fits.}
\item{de Wit: review, critique, and contextualize.}
\item{\st{Heun: Add option for $m = 1$ in single-factor models.}}
\item{\st{Heun: Re-implement confidence interval calculations to use
resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Include scatter graphs for model parameters 
estimated by resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Decide if the single factor models should include constraints
on $\alpha$, $\beta$, and $\gamma$. At present, the single-factor model allows
values outside the ranges $\alpha \in (0,1)$, $\beta \in (0,1)$, and $\gamma \in (0,1)$.}}
\item{\st{Heun: Make smaller scatter plots for resample coefficients for those 
countries where $u$ is involved.}}
\item{\st{Heun: Add a table that shows the fraction of unconverged resample fits
for each model, energy type, and factor.}}
\item{\st{Heun: Get AIC calculations working again.}}
\item{\st{Pruim: Add base fit red dots to all cloud plots.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\end{itemize}

@@

\noindent Equation~\ref{eq:CES} is a CES production function with 
capital stock ($k$) and labor ($l$) factors of production. 
Equation~\ref{eq:CESe} augments Equation~\ref{eq:CES} with energy 
using a ($kl$)($e$) nesting structure, as is typical in the literature. 
Equation~\ref{eq:CES} is a degenerate form of Equation~\ref{eq:CESe} 
where $\delta = 1$ and $\rho = -1$. The elasticity of substitution 
between capital ($k$) and labor ($l$) is given by 
$\sigma_1 = \frac{1}{1+\rho_1}$, 
the elasticity of substitution between ($kl$) and ($e$) is given by 
$\sigma = \frac{1}{1+\rho}$, 
$\delta_1 \in [0, 1]$ describes the relative importance of capital ($k$)
and labor ($l$), 
$\delta \in [0, 1]$ describes the importance 
of ($kl$) relative to ($e$), 
and $\gamma$ is a fitting parameter, 
expected to be close to unity. 
As $\sigma \rightarrow \infty$ ($\rho \rightarrow -1$), 
($kl$) and ($e$) are perfect substitutes. 
As $\sigma \rightarrow 0$ ($\rho \rightarrow \infty$), 
($kl$) and ($e$) are perfect complements: 
no substitution is possible. Similarly,

@@

\noindent According to the LINEX model, an economy that increases 
capital stock ($k$) without a commensurate increase in labor ($l$) 
and energy ($e$) will experience an increase in output ($y$) in excess 
of its increase of energy consumption ($e$), because $A$ will be 
greater than 1. Similarly, an economy benefits by increasing labor ($l$) 
without a commensurate increase in energy ($e$). Thus, $\rho_k$ is an 
efficiency of using additional labor and energy to make 
additional capital stock available to the economy, 
and $\rho_l$ is an efficiency of using additional energy to make 
additional labor available to the economy. 
$a_0$ represents the overall importance of these efficiencies, 
and $c_t$ represents the relative importance 
of the labor efficiency term compared to the capital stock efficiency term. 
We treat $a_0$ and $c_t$ as fitting parameters; 
both are assumed constant with respect to time.

@@

\noindent to ensure $\alpha + \beta = 1$ for constant returns to scale. 
$\lambda$ and $\alpha$ were estimated by the 
\texttt{R} \citep{R} function \texttt{nls}. 
If the estimated value for $\alpha$ was found outside the interval $[0,1]$, 
we set $\alpha$ to its boundary value and re-estimated $\lambda$. 
The value of $\beta$ was found with $\beta = 1 - \alpha$.

@@

Converting the resample data into 95\% confidence intervals 
provides data for Figure~\ref{fig:CD_Params_Graph}.
Appendix~\ref{sec:Model_Parameters} 
contains tables with numerical values of 
Cobb-Douglas model parameters and 95\% confidence intervals.

@@

\begin{equation} \label{eq:CDe_AB_reparam}
  y = A k^{\min(a,b)} l^{|b-a|} e^{1 - \max(a,b)}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent with 
$\alpha = \min(a,b)$, 
$\beta = |b-a|$, and 
$\gamma = 1 - \max(a,b)$.
The reparameterization ensures that $\alpha + \beta + \gamma = 1$. 
The \texttt{R} \citep{R} function \texttt{nls} was used to estimate 
values of $\lambda$, $a$, and $b$.
If the fitted value for $a$ ($b$) fell outside $[0,1]$, it was 
set equal to its boundary value and $\lambda$ and $b$ ($a$) 
were re-estimated.

To demonstrate the characteristics of the parameter estimation process, 
we performed a grid search over values of $\gamma$, 
wherein we re-estimated values of $\lambda$, $\alpha$, and 
$\beta$ at each value of $\gamma$. 
Figure~\ref{fig:CDeGridLattice} shows the sum of squared errors ($SSE$), 
$\alpha$, and $\beta$ as a function of the fixed value of 
$\gamma$ with thermal energy ($q$).

**** Fix the right y axis label. SSE should be italicized ($SSE$). 
Also, scaling on the axis labels and scales is not being picked up. ****

@@

Figure~\ref{fig:CDeGridLattice} shows that, for most economies,
the minimum $SSE$ occurs at or near a boundary of 
$\alpha$, $\beta$, or $\gamma$. 
The parameter estimation algorithm (\texttt{nls}) had difficulty converging when
the optimal value for the parameter represented by the absolute value function 
(in Equation~\ref{eq:CDe_AB_reparam}, $\beta$) 
was near a boundary. If the absolute value caused non-convergence of the 
\texttt{nls} function, second and, if necessary, third reparameterizations 
were attempted:

\begin{equation} \label{eq:CDe_CD_reparam}
  y = A k^{1 - \max(c,d)} l^{\min(c,d)} e^{|d-c|}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\begin{equation} \label{eq:CDe_FG_reparam}
  y = A k^{|g-f|} l^{1 - \max(f,g)} e^{\min(f,g)}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent If none of the three reparameterizations converged, 
the parameterization with the smallest $SSE$
was used. \ref{sec:Convergence_Resample_Models} shows the 
fraction of unconverged models for each model type. The above approach
resulted in very few unconverged resample models, 
even with \Sexpr{numResamples()} resamples.

Figure~\ref{fig:CDq_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ 
parameter values for Cobb-Douglas (with $q$) resample models.

@@

<<CES_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType=NA)
@

<<CES_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType=NA)
@

@@

<<CES_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\rho$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType=NA)
@

@@

<<CESq_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="Q")
@

<<CESq_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="Q")
@

<<CESq_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="Q")
@

<<CESx_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="X")
@

<<CESx_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="X")
@

<<CESx_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="X")
@

<<CESu_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="U")
@

<<CESu_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="U")
@

<<CESu_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="U")
@
